---
title: "Getting_trees_and_data"
author: "Claire C Winfrey"
date: "5/4/2020"
output: html_document
---
***NOTE: I had a plethora of issues when trying to use rotl beginning at chunk 3. It worked for a bit, but then I developed HTTP failure issues. I tried restating R, updating R to Arbor Day, making a new script, and updating rotl (no update was availble), and following the exact instructions in the "How to use rotl" vignette, all to no avail. 

****So, I instead went to DRYAD and found a phylogeny for leafflower moths.

Making a function to extract useful ott.id from Open Tree of Life and doing so for ant genus Formica.
```{r, eval = TRUE, echo=TRUE}
GetTreeFromOpenTree <- function(taxa) {
  library(rotl)
  library(ape)
  ott.id <- tnrs_match_names(taxa)$ott_id
  sub.tree <-tol_subtree(ott.id)
  import.info <- list(class(sub.tree), ott.id)
  return(import.info)
}

formica.rotl <- GetTreeFromOpenTree(taxa="Formica")
formica.rotl 
```
Now get Open Tree's current best estimate of the phylogeny for the group. They call this the tree of life; we can get the subtree for just this group.
```{r, echo = TRUE}
formica.tree <- tol_subtree(formica.rotl[[2]])
plot.phylo(formica.tree, type="fan", cex=0.2)
```
By looking at the ratio of the number of internal nodes to number of possible nodes, we can figure out how resolved a tree is! (subtracting 1 from each to account for the root node that must always exist).

***I tried to debug the code in the chunk below, pasted from the course website, by running individual lines in a regular R script. Unfortunately, the code returned a lot of errors that I was unable to fix. Most telling I think is that line 42: "formica.studies.ids returns "character(0)", and the code on line 45 "formica.study1.metadata <- get_study_meta(formica.studies)" returns an error of "Error: HTTP failure: 400 <html><body><h1>invalid request</h1></body></html>". So it seems perhaps something is wrong with the website?

```{r, eval=TRUE, error=TRUE}
	print(paste("The formica tree has ", Ntip(formica.tree), " terminals and ",
	Nnode(formica.tree), " internal nodes out of ",Ntip(formica.tree)-2,
	" possible, which means it is ",
	round(100*(Nnode(formica.tree)-1)/(Ntip(formica.tree)-3), 2),
	"% resolved", sep=""))

	# Open Tree can also return the original studies with the source trees.
	formica.studies <- studies_find_studies(property="ot:focalCladeOTTTaxonName",
	value="formica")
	formica.studies.ids <- unlist(formica.studies$study_ids)

	# Let's get info on the first study
	formica.study1.metadata <- get_study_meta(formica.studies[[1]][[1]])
	print(get_publication(formica.study1.metadata))

	# And let's get the tree from this study
	# Look in rotl documentation for the right function
	# Hint: get_study_trees() is close, but you don't know the tree.id
  formica.study1.tree1 <- get_tree_ids(formica.studies.ids[[1]])

	# And plot it
	plot.phylo(formica.study1.tree1, type="fan", cex=0.2)

	#Return both trees
	return.list <- list(formica.tree, formica.study1.tree1)
	names(return.list) <- c("tol.tree", "study1.tree")
	return(return.list)
}
```

HOMEWORK-- GET YOUR OWN DATA, LOAD IT IN, AND CHECK IT OUT BY PLOTTING IT AND SUMMARIZING IT

1) Errors using my own function (that worked above)
```{r, error =TRUE}

#I attempted with my own function:
phan.id <-GetTreeFromOpenTree(taxa= "Phanaeus")

```
And errors following ROTL guide:
```{r, error= TRUE}
library(rotl)
taxa <-c("Phanaeus", "Coprophanaeus", "Onthophagus", "Aphodius")
resolved_names <- tnrs_match_names(taxa, context_name = "Animals")
my_tree <-tol_induced_subtree(ott_ids = resolved_names$ott_id)
```
Because Open Tree of Life seems not to be working, I instead went to Dryad to find a phylogeny.

Citation: Luo, Shi-Xiao et al. (2016), Data from: A novel, enigmatic basal leafflower moth lineage pollinating a derived leafflower host illustrates the dynamics of host shifts, partner replacement, and apparent coadaptation in intimate mutualisms, Dryad, Dataset, https://doi.org/10.5061/dryad.t9140

I downloaded the phylogeny (which was made in RAxML), and then opened it in Atom to make sure that the file looked okay.

Reading in Newick file:
```{r, eval =TRUE, error = TRUE}
library(phytools)
leafflower_moth_tree <- read.newick(file = "/Users/clairewinfrey/Desktop/Spring_2020/Phylometh_2020/RAxML_Epicephala_result.txt")
leafflower_moth_tree
branching.times(leafflower_moth_tree)
```

Data looks good in text editor, is rooted, and has the taxon names that I expect!
```{r}
library(ape)
plot.phylo(leafflower_moth_tree, type = "fan", cex = 0.3)
```